---
title: "Reprocudible_Resarch_Week4"
author: "spoltone"
date: "29 09 2020"
output:
  word_document: default
  html_document:
    df_print: paged
---

# Data load
```{r}
library(knitr)
library(R.utils)
```

```{r}
raw_data <- read.csv(file = "repdata_data_StormData.csv", header = TRUE, sep = ",")

# summary

summary(raw_data)

# check dataset

head(raw_data)
names(raw_data)

```

## RESULTS

#### QUESTION 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

##### Reducing the data set to only needed columns

```{r}
main <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")

main_data<-raw_data[main]

names(main_data)
```

##### Reviewing events that cause the most fatalities and the most injuries

```{r}
Fatalities <- aggregate(FATALITIES ~ EVTYPE, data = main_data, FUN = sum)
Top10_Fatalities <- Fatalities[order(-Fatalities$FATALITIES), ][1:10, ] 
Top10_Fatalities 

Injuries <- aggregate(INJURIES ~ EVTYPE, data = main_data, FUN = sum)
Top10_Injuries <- Injuries[order(-Injuries$INJURIES), ][1:10, ] 
Top10_Injuries 
```

## Plot of Top 10 Fatalities & Injuries for Weather Event Types
```{r}
par(mfrow=c(1,2),mar=c(10,3,3,2))
barplot(Top10_Fatalities$FATALITIES,names.arg=Top10_Fatalities$EVTYPE,las=2,col="gray",ylab="fatalities",main="Top 10 fatalities")
barplot(Top10_Injuries$INJURIES,names.arg=Top10_Injuries$EVTYPE,las=2,col="gray",ylab="injuries",main="Top 10 Injuries")
```

Figure 1: The weather event responsbile for the highest fatalities and injuries is the ‘Tornado’

### QUESTION 2. Across the United States, which types of events have the greatest economic consequences?

# Q2.1 Data Exploration & Findings ...

# Upon reviewing the column names, the property damage(PROPDMG) and crop damage(CROPDMG) columns both have another related column titled 'exponents' (i.e - PROPDMGEXP and CROPDMGEXP respectively). 
# As a result, let's convert the exponent columns into numeric data for the calculation of total property and crop damages encountered.

```{r}
unique(main_data$PROPDMGEXP)
```

```{r}
# Assigning values for the property exponent main_data 
main_data$PROPEXP[main_data$PROPDMGEXP == "K"] <- 1000
main_data$PROPEXP[main_data$PROPDMGEXP == "M"] <- 1e+06
main_data$PROPEXP[main_data$PROPDMGEXP == ""] <- 1
main_data$PROPEXP[main_data$PROPDMGEXP == "B"] <- 1e+09
main_data$PROPEXP[main_data$PROPDMGEXP == "m"] <- 1e+06
main_data$PROPEXP[main_data$PROPDMGEXP == "0"] <- 1
main_data$PROPEXP[main_data$PROPDMGEXP == "5"] <- 1e+05
main_data$PROPEXP[main_data$PROPDMGEXP == "6"] <- 1e+06
main_data$PROPEXP[main_data$PROPDMGEXP == "4"] <- 10000
main_data$PROPEXP[main_data$PROPDMGEXP == "2"] <- 100
main_data$PROPEXP[main_data$PROPDMGEXP == "3"] <- 1000
main_data$PROPEXP[main_data$PROPDMGEXP == "h"] <- 100
main_data$PROPEXP[main_data$PROPDMGEXP == "7"] <- 1e+07
main_data$PROPEXP[main_data$PROPDMGEXP == "H"] <- 100
main_data$PROPEXP[main_data$PROPDMGEXP == "1"] <- 10
main_data$PROPEXP[main_data$PROPDMGEXP == "8"] <- 1e+08

# Assigning '0' to invalid exponent main_data
main_data$PROPEXP[main_data$PROPDMGEXP == "+"] <- 0
main_data$PROPEXP[main_data$PROPDMGEXP == "-"] <- 0
main_data$PROPEXP[main_data$PROPDMGEXP == "?"] <- 0
```

##### aggregate the property damage by the event type and sort the output it in descending order

```{r}
# property damage summary
propdmg <- aggregate(PROPDMG~EVTYPE,data=main_data,FUN=sum,na.rm=TRUE)
propdmg <- propdmg[with(propdmg,order(-PROPDMG)),]
propdmg <- head(propdmg,10)
print(propdmg)

# crop damage summary
crop <- aggregate(CROPDMG~EVTYPE,data=main_data,FUN=sum,na.rm=TRUE)
crop <- crop[with(crop,order(-CROPDMG)),]
crop <- head(crop,10)
print(crop)
```

# Plot of Top 10 Property & Crop damages by Weather Event Types

##plot the graph showing the top 10 property and crop damages

```{r}
par(mfrow=c(1,2),mar=c(11,3,3,2))
barplot(propdmg$PROPDMG/(10^9),names.arg=propdmg$EVTYPE,las=2,col="red",ylab="Prop.damage(billions)",main="Top10 Prop.Damages")
barplot(crop$CROPDMG/(10^9),names.arg=crop$EVTYPE,las=2,col="red",ylab="Crop damage(billions)",main="Top10 Crop.Damages")
```


